---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Jinyang Xu"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **[Minnesota] Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)

library(tidyverse)
library(dplyr)
library(knitr)
library(tidycensus)

# Set your Census API key

census_api_key("fef2af5588030f3fd4b474bd111e0d63a7ec2837")

# Choose your state for analysis - assign it to a variable called my_state

my_state  <- "MN"

```
**State Selection:** I have chosen **[Minnesota]** for this analysis because: It offers a strong balance of urban and rural diversity, allowing exploration of data quality across different community types.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.


```{r county-data}
# Write your get_acs() code here
county_data <- get_acs(
  geography = "county",
  state = my_state,
  variables = c(
    median_income = "B19013_001",   # Median Household Income
    total_population = "B01003_001" # Total Population
  ),
  year = 2022,
  survey = "acs5",
  output = "wide"
)

county_data
# Clean the county names to remove state name and "County" 

# Hint: use mutate() with str_remove()
county_data_clean <- county_data %>%
  mutate(
    NAME = str_remove(NAME, ",.*"),         # remove state name after the comma
    NAME = str_remove(NAME, " County$")     # remove trailing "County"
  )
# Display the first few rows
head(county_data_clean)
```


## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Create reliability categories
# Calculate MOE percentage and reliability categories using mutate()
county_data_clean <- county_data_clean %>%
  mutate(MOE = (median_incomeM / median_incomeE) * 100)

county_data_clean <- county_data_clean %>%
  mutate(reliability = case_when(
    MOE < 5 ~ "High Confidence",
    MOE >= 5 & MOE <= 10 ~ "Moderate Confidence",
    MOE > 10 ~ "Low Confidence"
  ))


county_data_clean <- county_data_clean %>%
  mutate(unreliable_flag = MOE > 10)
county_data_clean
# Create a summary showing count of counties in each reliability category

reliability_summary <- county_data_clean %>%
  count(reliability) %>%
  mutate(percentage = (n / sum(n)) * 100)

reliability_summary
# Hint: use count() and mutate() to add percentages
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
top5 <- county_data_clean %>%
  arrange(desc(MOE)) %>%
  slice(1:5) %>%
  select(
    County = NAME,
    Median_Income = median_incomeE,
    Margin_of_Error = median_incomeM,
    MOE_Percentage = MOE,
    Reliability = reliability
  )
# Format as table with kable() - include appropriate column names and caption
kable(top5, caption = "Top 5 Counties with Highest MOE Percentage (Median Income)")
```

**Data Quality Commentary:**

[High MOE means higher reliability for decision making. Small sample size or hidden factors account for high MOEs.]

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties

high_conf <- county_data_clean %>%
  filter(reliability == "High Confidence") %>%
  slice(2)

moderate_conf <- county_data_clean %>%
  filter(reliability == "Moderate Confidence") %>%
  slice(9)

low_conf <- county_data_clean %>%
  filter(reliability == "Low Confidence") %>%
  slice(1)

selected_counties_allv <- bind_rows(high_conf, moderate_conf, low_conf)

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties <- selected_counties_allv %>%
  select(NAME, median_incomeE, MOE, reliability, total_populationE)
selected_counties
```

**Comment on the output:** [The output is clear, but I don't know if it's good to ignore the GEOID :-]

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names

race <- c(
  white = "B03002_003",
  black = "B03002_004",
  hispanic = "B03002_012",
  total_pop = "B03002_001"
)

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter

selected_county_codes <- selected_counties_allv %>%
  mutate(county_code = str_sub(GEOID, 3, 5)) %>%  
  pull(county_code)

tract_data <- get_acs(
  geography = "tract",
  state = "27",
  county = selected_county_codes,
  variables = race,
  year = 2022,
  survey = "acs5",
  output = "wide"
)

tract_data

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
tract_data2 <- tract_data %>%
  mutate(pct_white = (whiteE / total_popE) * 100) %>%
  mutate(pct_black = (blackE / total_popE) * 100) %>%
  mutate(pct_hispanic = (hispanicE / total_popE) * 100)

tract_data2 

# Add readable tract and county name columns using str_extract() or similar

tract_data3 <- tract_data2 %>%
  mutate(
    tract_id = str_extract(NAME, "Census Tract \\d+"),     
    tract_id = str_remove(tract_id, "Census Tract "),
    
    county_name = str_extract(NAME, "(?<=; ).* County"),     
    county_name = str_remove(county_name, " County$"),
    
    state_name  = str_extract(NAME, "(?<=; )[^;]+$")         
  ) %>%
  relocate(tract_id, county_name, state_name, .after = NAME)

tract_data3
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract

top_hispanic <- tract_data3 %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) 

top_hispanic2 <- top_hispanic %>%
  select(tract_id,county_name,state_name,total_popM,pct_white,pct_black,pct_hispanic)

kable(
  top_hispanic2, 
  caption = "Tract with Highest Hispanic/Latino Percentage"
) 

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
county_summary <- tract_data3 %>%
  group_by(county_name) %>%
  summarise(
    n_tracts = n(),
    avg_white = mean(pct_white, na.rm = TRUE),
    avg_black = mean(pct_black, na.rm = TRUE),
    avg_hispanic = mean(pct_hispanic, na.rm = TRUE)
  ) %>%
  ungroup()

# Create a nicely formatted table of your results using kable()
kable(
  county_summary,
  digits = 1,  
  caption = "Average Demographics by County"
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
tract_moe <- tract_data3 %>%
  mutate(
    white_moe_pct    = (whiteM / whiteE) * 100,
    black_moe_pct    = (blackM / blackE) * 100,
    hispanic_moe_pct = (hispanicM / hispanicE) * 100
  )
tract_moe
# Hint: use the same formula as before (margin/estimate * 100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

tract_moe <- tract_moe %>%
  mutate(
    high_moe_flag = ifelse(
      white_moe_pct > 10 | black_moe_pct > 10 | hispanic_moe_pct > 10,
      TRUE, FALSE
    )
  )

# Create summary statistics showing how many tracts have data quality issues
moe_summary <- tract_moe %>%
  summarise(
    total_tracts = n(),
    tracts_high_moe = sum(high_moe_flag, na.rm = TRUE),
    pct_high_moe = (tracts_high_moe / total_tracts) * 100
  )

moe_summary
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison\

pattern_sum <- tract_moe %>%
  group_by(high_moe_flag) %>%
  summarise(
    n_tracts       = n(),
    avg_pop = mean(total_popE, na.rm = TRUE),
    avg_white_pct  = mean(pct_white, na.rm = TRUE),
    avg_black_pct  = mean(pct_black, na.rm = TRUE),
    avg_hispanic_pct = mean(pct_hispanic, na.rm = TRUE)
  ) %>%
  ungroup()


# Create a professional table showing the patterns

kable(
  pattern_sum,
  digits = 2,
  caption = "Comparison of Tracts by High vs. Low MOE"
)

# INCIDENT： The result showed that 100% of tracts were flagged as TRUE.
# Therefore, the next step is to take a more detailed approach

tract_moe1 <- tract_moe %>%
  mutate(
    white_high_moe    = white_moe_pct > 10,
    black_high_moe    = black_moe_pct > 10,
    hispanic_high_moe = hispanic_moe_pct > 10
  )


race_moe_sum <- tract_moe1 %>%
  summarise(
    white_flagged    = sum(white_high_moe, na.rm = TRUE),
    black_flagged    = sum(black_high_moe, na.rm = TRUE),
    hispanic_flagged = sum(hispanic_high_moe, na.rm = TRUE)
  )

race_moe_sum

```

**Pattern Analysis:** The analysis reveals a clear pattern: racial and ethnic minority populations have consistently less reliable data at the tract level. Specifically, the MOE exceeded 10% for all Black and Hispanic/Latino population estimates across the selected tracts, while White population estimates were somewhat more reliable, though still flagged in over half of the tracts.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

*Overall Pattern Identification*
The multi-level analysis — spanning county-level median income estimates and tract-level demographic breakdowns — reveals systematic data reliability challenges in ACS 5-year estimates
County-level: At county scale, median household income estimates show notable variability. Using a 10% MOE cutoff, most counties fell into moderate or low confidence categories, highlighting that uncertainty exists even at higher aggregation levels
Tract-level: Reliability issues intensify at the tract scale. Every tract exceeded the 10% MOE threshold for at least one demographic subgroup, 100% of tracts were flagged as high-MOE:(

*Equity Assessment*
The communities at greatest risk of algorithmic bias are:
Smaller or rural tracts: Populations with fewer respondents also face higher sampling error, compounding reliability problems.
Black and Hispanic/Latino communities: Their tract-level ACS estimates never meet the 10% reliability threshold. If an allocation algorithm treats these noisy estimates as precise, it will undervalue their needs and direct resources elsewhere.

*Root Cause Analysis*
The patterns are rooted in structural features of ACS data collection:
Sampling limits: Smaller population groups produce larger standard errors.
Differential nonresponse: Minority and immigrant communities may less likely to respond due to barriers and distrust.
Survey design constraints: The ACS 5-year estimates aggregate limited samples, and subgroup breakdowns at tract scale exceed the data’s design precision.

*Strategic Recommendations*
Data use：
Flag any tract with MOE% > 10 and avoid treating estimates as precise since sample size is small in tract level.
Use fallback geography (county) for variables with extreme MOE.

Strengthen inputs：
Blend multiple indicators (poverty, rent burden, child poverty) to reduce reliance on a single noisy variable.
Improve ACS participation via targeted outreach and language support in undercounted communities.

Outside the data:
Introduce human review for flagged tracts before final funding decisions.
Pilot and monitor the algorithm annually, recalibrating thresholds as survey data quality evolves.
Document and communicate uncertainty explicitly to policymakers and stakeholders.

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

sum_table <- county_data_clean %>%
  select(NAME, median_incomeE, MOE, reliability) %>%
  mutate(
    recommendation = case_when(
      reliability == "High Confidence" ~ "Safe for algorithmic decisions",
      reliability == "Moderate Confidence" ~ "Use with caution - monitor outcomes",
      reliability == "Low Confidence" ~ "Requires manual review or additional data"
    )
  )

# Format as a professional table with kable()

kable(
  sum_table,
  digits = 1,
  caption = "Decision Framework for Algorithmic Implementation"
)

```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** 
Aitkin, Anoka, Becker, Blue Earth, Brown, Carver, Cass, Chisago, Clay, Crow Wing, Dakota, Dodge, Fillmore, Freeborn, Goodhue, Hennepin, Hubbard, Isanti, Itasca, Kanabec, Kandiyohi, Lake, Le Sueur, Lyon, Marshall, Meeker, Mille Lacs, Morrison, Olmsted, Otter Tail, Pine, Pope, Ramsey, Renville, Rice, Roseau, St. Louis, Scott, Sherburne, Sibley, Stearns, Steele, Wabasha, Wadena, Washington, Wright

Counties listed above have suitable sample size and stronger ACS samples, which means lower MOE.
Demographic estimates are robust enough to support algorithm-driven funding without significant reliability concerns.
Algorithms here can be implemented immediately with standard fairness monitoring.

2. **Counties requiring additional oversight:** 
Beltrami, Benton, Big Stone, Carlton, Chippewa, Clearwater, Cook, Cottonwood, Douglas, Faribault, Grant, Houston, Jackson, Kittson, Koochiching, Lac qui Parle, Lincoln, McLeod, Mahnomen, Martin, Mower, Murray, Nicollet, Nobles, Norman, Pennington, Pipestone, Polk, Red Lake, Redwood, Rock, Swift, Todd, Waseca, Watonwan, Wilkin, Winona, Yellow Medicine.

Recommended monitoring actions for counties above:
Outcome audits: Compare algorithmic allocations against program participation data.
Sensitivity testing: Run models with upper/lower confidence bounds (estimate ± MOE) to test robustness.
Flagging system: Automatically flag tracts/counties with borderline estimates for human review.

3. **Counties needing alternative approaches:**ss
Traverse

Suggested alternatives:
Manual review: Use staff expertise and local knowledge to adjust allocations.
Supplemental data: Incorporate administrative records (school eligibility, Medicaid enrollment, unemployment claims) to cross-validate needs.
Community engagement: Consult local service providers or conduct targeted surveys to supplement missing precision.

## Questions for Further Investigation
Beyond race/ethnicity, do other vulnerable groups (e.g., low-income households, elderly residents, non-English speakers) also experience systematically higher MOE?
Could combining ACS with administrative datasets (SNAP, TANF, Medicaid) reduce disparities in data reliability?
How stable are tract-level estimates over time, and do reliability issues persist across survey cycles?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [2025/09/27]

**Reproducibility:** 
- All analysis conducted in R version [4.5.1]
- Census API key required for replication
- Complete code and documentation available at: [portfolio-setup-jyxu48/Assignments/assignment_1/]

**Methodology Notes:**
County Selection: Minnesota was chosen as the state of analysis, with Hennepin County initially used as a representative large metropolitan area. The analysis was later expanded to include all counties statewide to examine systematic patterns.

Variable Choice: We relied on ACS 5-year (2022) estimates for median household income, total population, and racial/ethnic composition (White alone, Black/African American alone, and Hispanic/Latino). Results therefore depend on ACS methodology and survey design.

Reliability Thresholds: Data quality was assessed by calculating MOE percentages (MOE ÷ estimate × 100). For this project, a 10% cutoff was applied to flag “high MOE,” with <5% labeled as High Confidence, 5–10% as Moderate Confidence, and >10% as Low Confidence. The 10% setting is more stringent than some standard practices (e.g., 15%), which increases sensitivity to data reliability concerns.

**Limitations:**
ACS tract-level estimates have sampling errors, especially for small subgroups.
Findings sensitive to thresholds, year, and variable selection.
Results reflect ACS survey design, not absolute ground truth.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [·] All code chunks run without errors
- [·] All "[Fill this in]" prompts have been completed
- [·] Tables are properly formatted and readable
- [·] Executive summary addresses all four required components
- [·] Portfolio navigation includes this assignment
- [·] Census API key is properly set 
- [·] Document renders correctly to HTMLs

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`